Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-04-21 main.tex
Wordsmithing the abstract and introduction.

\abstract{
  Uncontrolled bicycles are unstable at low speeds. We add a controlled
  steering motor to a consumer bicycle that stabilizes the bicycle at low
  speeds and apply varying magnitude external handlebar perturbations to the
  bicycle while ridden on a treadmill. The probability of not recovering from a
  handlebar perturbation decreases when the balance assist is activated.
}

\section{Introduction}
%
Single actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and are challenging to
control for the rider. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds. Bicycles can be mechanically modified to lower the
speeds at which it is self-stable~\citep{Astrom2005} but since at least the
1980s it is known that applying a steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If the robotic control
of steering can stabilize a bicycle, it may reduce the control required by the
rider to successfully manage balancing tasks. We have developed such a balance
assisting bicycle and hypothesize that it helps the rider in situations they
are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined throughout the 1970s.
\Citet{vanZytveld1975} seems to be the first
\todo{Check if a Japanese researcher was first}
to attempt to robotically stabilize a small motorbike with a controlled
inverted pendulum that mimicked rider lean, but he was not successful in
demonstrating what his control model predicted. In his model, he recognized
that feedback of the vehicle roll angle and angular rate was essential to
stabilize the vehicle. It was not until the mid 1980s that someone successfully
demonstrated an automatically balanced motorcycle~\citep{Ruijs1986a} and they
did so with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, roll angular
rate feedback stabilizes the weave mode, and steer angular rate feedback
stabilizes the wobble mode. They also show how the gains need to change with
respect to vehicle speed for favorable control at all speeds. This roll motion
feedback is the simplest controller that can stabilize a single track vehicle
above a minimum speed when not concerned with wobble instabilities,
which are not prominent in bicycles. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved.  Many more automatically balanced
single track vehicles have been demonstrated over the last 40 years, but none
have demonstrated that increasing low speed stability can assist in balancing
and possibly reduce single actor falls. Most of these robotic bicycles and
motorcycles designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel. The effect of this automatic
control gives the ability to change the dynamics, up to some limits, of the
human controlled plant.

The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Interestingly, Whipple showed that roll motion feedback
stabilizes the system back in 1899 and the model is ideal for showing the
effect of a roll motion feedback driven steer motor on the dynamics. The linear
model can be described by the fourth order state space equations:

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add a
  controlled steering motor to a consumer bicycle that stabilizes the bicycle
  at low speeds. To test the motor's assistance during falls, we apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill with the balance assist system activated and deactivated.  The
  probability of not recovering from a handlebar perturbation decreases when
  the balance assist is activated at a speed of 6~\si{\kph}.
}

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds, possibly relieving the rider of a necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\citep{Astrom2005} but since at least the 1980s it
is known that applying a steering torque proportional to the roll rate can
stabilize a single track vehicle at very low speeds. If the robotic control of
steering that implements this can stabilize a bicycle, it may reduce the
control required by the rider to successfully manage balancing tasks. We have
developed such a balance assisting bicycle and hypothesize that it helps the
rider in situations they are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
researcher was first} to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
someone successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986a} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains need to change with respect to vehicle speed for
favorable control at all speeds. This roll motion feedback is the simplest
controller that can stabilize a single track vehicle above a minimum speed when
not concerned with wobble instabilities, which are not prominent in bicycles at
low speeds. Ruijs and Pacejka's work was not particularly concerned with low
speed stability and their vehicle was fully automatic, no human rider was
involved.  Many more automatically balanced single track vehicles have been
demonstrated over the last 40 years, but none have demonstrated that increasing
low speed stability can assist in balancing and possibly reduce single actor
falls. Most of these robotic bicycles and motorcycles designers did not intend
for a human rider to also control the stabilized vehicle. Nevertheless, an
automatically stabilized bicycle can be controlled by a human rider if the
motor controlled steer torque and the rider applied steer torque act on the
steer in parallel. The effect of this automatic control gives the ability to
change the dynamics, up to some limits, of the human controlled plant.

The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Whipple showed that roll motion feedback can stabilize the
system. The model is ideal for showing the effect of a roll motion feedback
driven steer motor on the dynamics. The linear model can be described by the
fourth order state space equations:
